Visualizing Conditional Relationships

### 2017-08-30 paper/main.tex
more spellcheck

==\documentclass[letterpaper,onecolumn,titlepage]{Ythesis}

@@

\title{Visualizing Functional Data}
\author{Hannah Aizenman}
\committee{Dr. Michael Grossberg(Advisor), Dr. Robert Haralick, Dr. Huy Vo}
\submitted{}
\abstract{There are an almost overwhelming number of machine learning and
  visualization techniques, but each technique makes some underlying
  assumptions about how the data is structured. While sometimes the
  observations are completely independent from each other, in many instances
  the observations are sampled from 1D continuum like time or 2d surface such
  as space.This survey uses functional data analysis to classify visualizations
  based on the number of variables in and level of paramterization of the data.}
\begin{document}
\makefrontmatter

\subfile{sections/intro.tex}
\subfilt{sections/problem.tex}
\subfile{sections/independent.tex}

\section{Conclusion}
\label{sec:conclusion}

\begin{tabular}{|r|r|r|r|}
  \hline
              & univariate & bivariate & multivariate\\
  \hline
   0D         &            &           &             \\
  \hline
   1D         &            &           &             \\
   \hline
   ND         &            &           &             \\
\end{tabular}

\pagebreak
\printbibliography
\end{document}

>>>>

\documentclass[letterpaper,onecolumn,titlepage]{Ythesis}

@@

\title{Visualizing Conditional Relationships}
\author{Hannah Aizenman}
\committee{Dr. Michael Grossberg(Advisor), Dr. Robert Haralick, Dr. Huy Vo}
\submitted{}
\abstract{The context surrounding data is often as important as the
  measurements themselves. Income data provides a lens into demographic
  inequalities when separated by race and gender; temperature data indicates
  seasons when plotted by time. These are relatively simple relationships, but
 many datasets are embedded with rather complex conditional relationships. This
 survey gives an overview of techniques showing conditional dependencies
 amongst distributions and densities. 
}
\begin{document}
\makefrontmatter

\section{Introduction}
\label{sec:introduction}
\begin{figure}
  \includegraphics{chart_chooser}
  \caption{From Bertin's Semiology of Graphics \cite{bertin}, this diagram
    illustrates how there are many ways to display the same information about the 1954 Frnech
  workforce.}
  \label{fig:chart_chooser}
\end{figure}

There are almost an overwhelming number of visualizations, and no real correct
choice for any given situation. As figure~\ref{fig:chart_chooser} shows, often
there isn't even a single standout chart for a single dataset; instead the
choice of figure is dependent on the relationships the researcher would like to
extract from the dataset. Is it important to know how many people are in each
department? A bar chart is suited for this, but this data has that information
segmented into sectors - should it be combined or stay separated? Is the
spatial distribution of the workforce something that should be preserved? If
so, how should the sector information be retained? Sectors and space can both
conditional attributes of the data, yielding information on how the workforce
varies when disaggregated into these smaller blocks.

Quite often, it is this disaggregation that is interesting to researchers
because it points to multiple factors that can contribute to the measured
outcome. The difficulty is in figuring out how to visualize this
disaggregation, especially as the data structure grows more complex. Cross
tabulation tables work for a small number of categorical variables, but are
pushed to the limits at n=3 or 4. And most data is not solely categorical,
instead often having quantitative densities of one sort or another. A complex but
important form of quantitative densities are functional densities, where they are drawn from a continuous sample of observations \cite{ramsey2006, ramsey2002, muller2006}. That continuous sample is often over space or time or another parameter or combination thereof, and often it is
important to preserve those functional dependencies in the visualization. For
example, for storm modeling it is important to show the expected time and
location of the rainfall. This survey presents a sampling of visualization
techniques that attempt to illustrate these conditional and functional
relationships within a dataset.

\begin{table}

\begin{tabular}
                & Distribution           & Density 
   Distribution & P(X|Y),P(A,B,C|E,F,G)  & 
   Density      &              & p(X|Y), p(A,B,C|E,F,G)
\end{tabular}

\caption{The rows represent the known distributions & densities while the
  columns are the distributions and densities that the visualization aims to derive.}
\end{table}

While this survey focuses on data structured in a linear, tabular, or
data cube way\cite{munzner data structure}, it could potentially be expanded to network data where the xi
parameters represent edges in the graph.

\subfile{sections/distdist.tex}
\subfile{sections/distdens.tex}
\subfile{sections/densdist.tex}
\subfile{sections/densdens.tex}

\section{Conclusion}
\label{sec:conclusion}
When trying to choose a visualization technique for a problem, it is often
crucial to interrogate the questions being asked of the data. What
relationships are important to preserve and which can be discarded? While there
are a what feels like infinite number of visualization techniques, this
visualization task can be rather complicated, especially as the number of
variables or dependencies increases. Interactive tools can help, but most are limited to showing pairwise or a small number of connected interactions. Machine learning can somewhat be used
to explore which variables and dimensions, and therefore which relationships, are worth
exploring further, but often the results of those algorithms need to
be translated to the visual space and those visualizations need to be
understandable.

\printbibliography
\end{document}

### 2017-08-30 paper/sections/densdens.tex
more spellcheck

While bar graphs and histograms can give asense of the variability in
data, they are limited to showing the dispersion of the data only in how the measurements are spread out. This is somewhat limiting for very large datasets, so the boxplot has evolved as a way to display more characteristics of the distribution \cite{wickham2011}. While a simple boxplot does not inherently show conditional dependencies, many derviations do.

\begin{figure}
  \includegraphics{boxplot}
  \caption{Tukey's 1977 example of the box plot (exhibit 6 of chapter 2 from
    Exploratory Data Analytics\cite{tukey1977}) illustrates how the box plot
    shows
    the distribution of topology of states and
    the distribution of volcanos. The strength of the box and whisker is that
    the
    reader can quickly compare the two distributions and learn, for example,
    that
    volcanos tend to span a shorter range of average heights, but that the
    extremes
    are much further from the center. Essentially, the distributions act as
    references for the other displayed distributions.}
  \label{fig:boxplot}
\end{figure}

Tukey's introduction to the boxplot \cite{tukey1977}, shown in
Figure~\ref{fig:boxplot} illustrates how the distributions of these 1d densities differ by category. While this graph shows the median, range between upper and lower interquartile, and lower and upper extreme, many authors have exploited the structure to convey more information about the data. Authors used different quantile levels \cite{hyndman1996}s or measures of outliers
\cite{frigge1989}, or ways of identifying outliers\cite{carter2009,
  schwertman2004, schwertman\
  2007}, or otherwise incorporated skewness,
kurtosis, and other descriptive distributional statistics \cite{kim2004,
  hubert2008, marmolejo\
  2015}.

>>>>

While bar graphs and histograms can give a sense of the variability in
data, they are limited to showing the dispersion of the data only in how the measurements are spread out. This is somewhat limiting for very large datasets, so the box plot has evolved as a way to display more characteristics of the distribution \cite{wickham2011}. While a simple box plot does not inherently show conditional dependencies, many derivations do.

\begin{figure}
  \includegraphics{boxplot}
  \caption{Tukey's 1977 example of the box plot (exhibit 6 of chapter 2 from
    Exploratory Data Analytics\cite{tukey1977}) illustrates how the box plot
    shows
    the distribution of topology of states and
    the distribution of volcanoes. The strength of the box and whisker is that
    the
    reader can quickly compare the two distributions and learn, for example,
    that
    volcanoes tend to span a shorter range of average heights, but that the
    extremes
    are much further from the center. Essentially, the distributions act as
    references for the other displayed distributions.}
  \label{fig:boxplot}
\end{figure}

Tukey's introduction to the box-plot \cite{tukey1977}, shown in
Figure~\ref{fig:boxplot} illustrates how the distributions of these 1 dimensional densities differ by category. While this graph shows the median, range between upper and lower interquartile, and lower and upper extreme, many authors have exploited the structure to convey more information about the data. Authors used different quantile levels \cite{hyndman1996}s or measures of outliers
\cite{frigge1989}, or ways of identifying outliers\cite{carter2009,
  schwertman2004, schwertman\
  2007}, or otherwise incorporated skewness,
kurtosis, and other descriptive distributional statistics \cite{kim2004,
  hubert2008, marmolejo\
  2015}.

While box plots were designed to visualize discrete observations, often there
are no discrete observations and instead the `observation` is a timeseries
curve (where the connections are as important as the individual points) or
spatial region. Understanding the distributions of functional observations
often relies first ordering the functions such that there's some measure of how
close these functions are to each other--functional depth\cite{febrero2007},
bivariate score depth\cite{hyndman2009}, and
bivariate kernel density estimation\cite{scott1992} are some examples. This
ordering then yields functional analogs to median and other percentile
metrics. For small values, the ordered curves can then be colored based on the
ordering metric \cite{hyndman2009}, but this method is unwieldy for larger
ensembles.

\begin{figure}
  \includegraphics[width=.75\linewidth]{contour_weather}
  \caption{From \cite{whitaker2013}, the contour box plot method is used to show uncertainty in temperature ensembles while maintaining the temporal dimension of the data.}
  \label{fig:countour}
\end{figure}
Functional bag, box, and HDR plots \cite{hyndman2009, sun2011}} tend to be good at visualizing specific types of outliers, but are somewhat limited to only one functional distribution. A very common task is to compare ensembles, each a function of the same timeseries or geographic region. Contour box plots\cite{whitaker2013} aim to capture the variability of the
uncertainty via a band depth ordering metric. Each ensemble members band depth
is computed as sum of the probabilities that the observations in any given
ensemble fall within the max-min envelope defined by any two other
ensembles. The bands are then sorted by band depth such that the median is the
ensemble with about 50\% of its members whiten all envelopes formed by other
bands (so most centered). Outer bands are chosen according to the task at
hand, which in this paper is temperature visualizations. The authors argue that contour box plots are an improvement over spaghetti plots \cite{lau_koa_pang} because the spaghetti plots get nosy as the number of plots increase and so it's hard to tease out specific patterns. Instead in
Figure~\ref{fig:countour}, the authors remove most of the bands and instead visualize an outlying envelope(light gray) and a more central envelope(dark gray). It retains much of the information of
the spaghetti, but removes the visual noise of the lines.

One of the most common approaches for visualizing datasets that are structurally complex or have more than one or two levels of conditional dependencies is to built a toolkit comprising of many different basic visualization types. The calendar visualization described above is a simple example, and just like more complex tools it follows the design principal of overview-zoom/filter-details-on-demand \cite{shneiderman1996eyes} because these tools facilitate engagement with the data at the aggregate and encourage drilling down to the core. 
\end{document}

### 2017-08-30 paper/sections/densdist.tex
more spellcheck

Many multivariate visualizations treat the data space as a graph wherin the vertices are the variables and the edge corresponds to some quantified relationship between the two vertices and often these edges are colored by category or form a shape that can be used a proxy for a category. These types of visualizations can often be used to derive typical observations in a given category.

\begin{figure}
  \includegraphics{flexaxis.png}
  \caption{Figure taken from \cite{claessen_van_wijk_2011} illustrates the following common
    methods for visualizing multivariate datasets: matrix of scatters, parallel coordinates and rader charts.
  \label{fig:flexlink}
\end{figure}

There are a number of methods for showing how measurments of different
variables interact, and as showing figure~\ref{fig:flexlink} color is often
used to illustrate how these relationships are different for each category of
data. As seen in figure~\ref{fig:flexlink}, the common methods for showing
visualiztions are matrix of scatters \cite{elmqvist, } , parallel
coordinates \cite{islenberg, wegman} and rader charts\cite{chambers_tukey}/

\begin{figure}
  \includegraphics{parallelsets.png}
  \caption{On the left is the categorical view, where users can define
    categories based on other categorical information. For example, in this
    veiw working family with children is defined as a household with one or
    more children. On the right is a parallel coordinate graph wherin the
    dominiant category informs the colors, the size of the box is relative to
    ghe category size, and the historgrams show the freqiencies of those
    categories relative to the connecting catgories (market type and income
    respectively). }
  \label{fig:parallelsets}
\end{figure}

Another important factor in visualizing categorical data is retaining the
inherent heirachy in many datasets with categorical measurements
\cite{shneiderman_visualizing_2000}. The Parallel Sets tool
showin in figure~\ref{fig:parallelsets} is expicilitly designed to show frequencies
of set membership in heirarichal ordered categorical sets, so the \{d_0, \dots, d_n} \in D
  \cite{kosara_parallel_2006}. Building on the fexible linked axis version of
  the parallel cooridnates plot \cite{claessen_flexible_2011}, Parallel Sets
  treats each categorical set independantly but indicates conditional
  dependency through cross tabulations and by grouping the data, indicated via
  color, based on an active dimension that then becomes the pivot point for the
  cross tabulations. In Parallel Sets, quantative data is converted to
  categorical via binning. Many-to-many PCP is another type of generalized
  parallel ccoordinates where all possible pairs are explored \cite{lind}.

\begin{figure}
  \includegraphics{corr.png}
  \label{caption} Zhang et al create an initial map (a) that they then apply an
  edge filter to (b). They then allow the user to filter out a variable and
  either highlight only the vertices of that variable (c) or the edges incident
  on that variable(d).
  \label{fig:corr}
\end{figure}

Often though it's important to retain the internel structure of the
quantative data, such as when it has internal structucture such as time and
space. Zhang et al propose a method that retains data and allows for
visualizing more than just pairwise relationships\cite{zhang_visual_2015}. They
introduce a concept of a correlation map, as shown in
figure~\ref{fig:corr}. The authors build out a graph where every variable is a
vertex and the edges are correlations between the connecting vertices. The
vertices are sized relative to the variance in the data. The edges are colored
red for negative correlations and green for positive correlations. To convert
the categorical data to numerical space, they developed an algorithm where when
mapped to numbers, the residual sum of squares of a logostic regression of the
categorical variables  would be minimized. They then filter edges with low
correlations and on variables of interest.
\end{document}

>>>>

Many multivariate visualizations treat the data space as a graph where the vertices's are the variables and the edge corresponds to some quantified relationship between the two vertexes and often these edges are colored by category or form a shape that can be used a proxy for a category. These types of visualizations can often be used to derive typical observations in a given category.

\begin{figure}
  \includegraphics{flexaxis.png}
  \caption{Figure taken from \cite{claessen_van_wijk_2011} illustrates the following common
    methods for visualizing multivariate datasets: matrix of scatters, parallel coordinates and radar charts.
  \label{fig:flexlink}
\end{figure}

There are a number of methods for showing how measurements of different
variables interact, and as showing figure~\ref{fig:flexlink} color is often
used to illustrate how these relationships are different for each category of
data. As seen in figure~\ref{fig:flexlink}, the common methods for showing
visualizations are matrix of scatters \cite{elmqvist, } , parallel
coordinates \cite{islenberg, wegman} and radar charts\cite{chambers_tukey}/

\begin{figure}
  \includegraphics{parallelsets.png}
  \caption{On the left is the categorical view, where users can define
    categories based on other categorical information. For example, in this
    view working family with children is defined as a household with one or
    more children. On the right is a parallel coordinate graph where the
    dominant category informs the colors, the size of the box is relative to
    the category size, and the histograms show the frequencies of those
    categories relative to the connecting categories (market type and income
    respectively)}
  \label{fig:parallelsets}
\end{figure}

Another important factor in visualizing categorical data is retaining the
inherent hierarchy in many datasets with categorical measurements
\cite{shneiderman_visualizing_2000}. The Parallel Sets tool
showin in figure~\ref{fig:parallelsets} is explicitly designed to show frequencies
of set membership in heirarichal ordered categorical sets, so the \{d_0, \dots, d_n} \in D
  \cite{kosara_parallel_2006}. Building on the flexible linked axis version of
  the parallel coordinates plot \cite{claessen_flexible_2011}, Parallel Sets
  treats each categorical set independently but indicates conditional
  dependency through cross tabulations and by grouping the data, indicated via
  color, based on an active dimension that then becomes the pivot point for the
  cross tabulations. In Parallel Sets, quantitative data is converted to
  categorical via binning. Many-to-many PCP is another type of generalized
  parallel coordinates where all possible pairs are explored \cite{lind}.

\begin{figure}
  \includegraphics{corr.png}
  \label{caption} Zhang et al create an initial map (a) that they then apply an
  edge filter to (b). They then allow the user to filter out a variable and
  either highlight only the vertexes of that variable (c) or the edges incident
  on that variable(d).
  \label{fig:corr}
\end{figure}

Often though it's important to retain the internal structure of the
quantitative data, such as when the data is a function of time and
space. Zhang et al propose a method that retains data and allows for
visualizing more than just pairwise relationships\cite{zhang_visual_2015}. They
introduce a concept of a correlation map, as shown in
figure~\ref{fig:corr}. The authors build out a graph where every variable is a
vertex and the edges are correlations between the connecting vertexes. The
vertexes are sized relative to the variance in the data. The edges are colored
red for negative correlations and green for positive correlations. To convert
the categorical data to numerical space, they developed an algorithm where when
mapped to numbers, the residual sum of squares of a logistic regression of the
categorical variables  would be minimized. They then filter edges with low
correlations and on variables of interest.
\end{document}
